A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database

### 2022-02-16 paper/paper.Rmd
conclusions

Why is this exercise important for teachers and students? The data package and software could be used for illustrating data cleaning in the classroom. The refreshed and contemporary data provided in the package can be used as examples in the teaching of initial data analysis and modeling, and exploratory data analysis of longitudinal data.

The wages data provides a good opportunity to discuss the difference between statistics to use for public policy and statistics that relate to the individual. From modeling the wages relative to workforce experience, with demographic covariates we learn that there are significantly different patterns. That more education leads to increasingly higher wages, which is a satisfying result for educators. It means that education makes a difference in the wage experience, and that public policy that encourages education is a data-supported action. We would also learn, although not presented here, that race matters and that being black leads to lower wages. This is a disturbing finding because there is no rational for such a difference in a fair society. This is would provide support for action in public policy to remove this overall average effect. It also provides an example for educators to explain the use of data to support public policy action.

These are interpretations of results from analyzing the wages data, which rely on trusting that the data provided is accurate and valid. The data is collected from a reputable organization, but the data has obvious errors that need correcting. This paper has illustrated procedures and guidelines to achieve valid data, and provides the code and details for it to be reproduced, and modified if deemed appropriate.

Having trustworthy data is imperative for statistics and data science education. Students do interpret results and take seriously what is reported from example analyses. If the data in the examples are synthetic, with our own inherent prejudices (on fog e.g. gender and race) applied, or inaccurately processed leaving errors, we as educators are being irresponsible because the societal message taught to students is flawed. This paper demonstrates the process to produce a trustworthy data set for teaching.

>>>>

Why is this exercise important for teachers and students? This work illustrates the steps in cleaning and processing data in the preparation for it to be a textbook data set. Choices made during the cleaning and processing can affect findings made with the data, and these should be transparent. Along with the textbook data, which is now provided as an R package, the documentation of the process provide some examples for teaching data cleaning, initial data analysis and exploratory data analysis. The refreshed textbook data provides a resource for teaching longitudinal data analysis.

The wages data provides a good opportunity to discuss the difference between statistics to use for public policy and statistics that relate to the individual. Public policy is based on models, yielding averages that might vary across strata. Modeling the wages relative to workforce experience, with demographic covariates we learn that there are significantly different patterns. That more education leads to increasingly higher wages, which is a satisfying result for educators. It means that education makes a difference in the wage experience, and that public policy that encourages education is a data-supported action. We would also learn, although not presented here, that race matters and that being black leads to lower wages. This is a disturbing finding because there is no rational for such a difference in a fair society. This would provide support for action in public policy to remove this overall average effect. It also provides an example for educators to explain the use of data to support public policy action.

On an individual level, one needs to know where am I in this data, and does this data relate to me. To do this, the individual profiles need to be explored. Pre-dominantly, we would learn that the variation from one individual to another is far more than the variation between demographic strata. For example, many individuals with lower educational attainment earn very high wages. From a statistics and data science educator perspective, more focus and more methodology for this type of statistics needs to be included in the curriculum.

The above interpretations of results from analyzing the wages data, rely on trusting that the data provided is accurate and valid. The wages data is collected by a reputable organization, but we found that the data has obvious errors that should be corrected. This paper has illustrated procedures and guidelines to achieve valid data, and provides the code and details for it to be reproduced, and modified if deemed appropriate.

Having trustworthy data is imperative for statistics and data science education. Whenever one uses a textbook data set that is perceived as relating to the students lives, there will be interpretations made. The wages data is an example of this. Students will take away the interpretations from whatever is taught with the data, that wages increase with experience, education, race. If one uses data examples that are synthetic, instilled with our own inherent prejudices (for e.g. gender and race), or data that has been poorly processed containing errors, we as educators are being irresponsible because the societal message taught to students may be flawed. This paper demonstrates the process to produce a trustworthy data set for teaching.

### 2022-02-16 paper/paper.tex
conclusions

Why is this exercise important for teachers and students? The data package and software could be used for illustrating data cleaning in the classroom. The refreshed and contemporary data provided in the package can be used as examples in the teaching of initial data analysis and modeling, and exploratory data analysis of longitudinal data.

The wages data provides a good opportunity to discuss the difference between statistics to use for public policy and statistics that relate to the individual. From modeling the wages relative to workforce experience, with demographic covariates we learn that there are significantly different patterns. That more education leads to increasingly higher wages, which is a satisfying result for educators. It means that education makes a difference in the wage experience, and that public policy that encourages education is a data-supported action. We would also learn, although not presented here, that race matters and that being black leads to lower wages. This is a disturbing finding because there is no rational for such a difference in a fair society. This is would provide support for action in public policy to remove this overall average effect. It also provides an example for educators to explain the use of data to support public policy action.

These are interpretations of results from analyzing the wages data, which rely on trusting that the data provided is accurate and valid. The data is collected from a reputable organization, but the data has obvious errors that need correcting. This paper has illustrated procedures and guidelines to achieve valid data, and provides the code and details for it to be reproduced, and modified if deemed appropriate.

Having trustworthy data is imperative for statistics and data science education. Students do interpret results and take seriously what is reported from example analyses. If the data in the examples are synthetic, with our own inherent prejudices (on fog e.g.~gender and race) applied, or inaccurately processed leaving errors, we as educators are being irresponsible because the societal message taught to students is flawed. This paper demonstrates the process to produce a trustworthy data set for teaching.

>>>>

Why is this exercise important for teachers and students? This work illustrates the steps in cleaning and processing data in the preparation for it to be a textbook data set. Choices made during the cleaning and processing can affect findings made with the data, and these should be transparent. Along with the textbook data, which is now provided as an R package, the documentation of the process provide some examples for teaching data cleaning, initial data analysis and exploratory data analysis. The refreshed textbook data provides a resource for teaching longitudinal data analysis.

The wages data provides a good opportunity to discuss the difference between statistics to use for public policy and statistics that relate to the individual. Public policy is based on models, yielding averages that might vary across strata. Modeling the wages relative to workforce experience, with demographic covariates we learn that there are significantly different patterns. That more education leads to increasingly higher wages, which is a satisfying result for educators. It means that education makes a difference in the wage experience, and that public policy that encourages education is a data-supported action. We would also learn, although not presented here, that race matters and that being black leads to lower wages. This is a disturbing finding because there is no rational for such a difference in a fair society. This would provide support for action in public policy to remove this overall average effect. It also provides an example for educators to explain the use of data to support public policy action.

On an individual level, one needs to know where am I in this data, and does this data relate to me. To do this, the individual profiles need to be explored. Pre-dominantly, we would learn that the variation from one individual to another is far more than the variation between demographic strata. For example, many individuals with lower educational attainment earn very high wages. From a statistics and data science educator perspective, more focus and more methodology for this type of statistics needs to be included in the curriculum.

The above interpretations of results from analyzing the wages data, rely on trusting that the data provided is accurate and valid. The wages data is collected by a reputable organization, but we found that the data has obvious errors that should be corrected. This paper has illustrated procedures and guidelines to achieve valid data, and provides the code and details for it to be reproduced, and modified if deemed appropriate.

Having trustworthy data is imperative for statistics and data science education. Whenever one uses a textbook data set that is perceived as relating to the students lives, there will be interpretations made. The wages data is an example of this. Students will take away the interpretations from whatever is taught with the data, that wages increase with experience, education, race. If one uses data examples that are synthetic, instilled with our own inherent prejudices (for e.g.~gender and race), or data that has been poorly processed containing errors, we as educators are being irresponsible because the societal message taught to students may be flawed. This paper demonstrates the process to produce a trustworthy data set for teaching.
